---
title: "A_2"
output: html_document
date: '2022-09-21'
---

Installing all required libraries and setting seed for reproducability
```{r}
# Attach the DESeq2 library
library(DESeq2)

# Attach the ggplot2 library for plotting
library(ggplot2)

# We will need this so we can use the pipe: %>%
library(magrittr)

# Tidyverse
library(tidyverse)

# Heat Map
library(pheatmap)

set.seed(12345)
```

Setting the path and getting dataframes
```{r}
setwd('/Users/sckaff/Desktop/CGS4144/Assignment_2/Data/')
data <- read.csv('data_csv.csv')
clinical <- read.csv('clinical_csv.csv')
```

Cleaning the data
```{r}
rownames(data) <- data$index
data$index = NULL
clinical <- clinical %>% rename(Condition = Primary.Cytogenetic.Code)
```

Initializing DESeq object
```{r}
ddset <- DESeqDataSetFromMatrix(
  countData = data,
  colData = clinical, 
  design = ~Condition
)
```

Creating DESeq object and storing its results
```{r echo=False}
deseq_object <- DESeq(ddset)
deseq_results <- results(deseq_object2)
```

Here we will use `lfcShrink()` function to obtain shrunken log fold change estimates based on negative binomial distribution.
This will add the estimates to your results table. 
Using `lfcShrink()` can help decrease noise and preserve large differences between groups (it requires that `apeglm` package be installed) [@Zhu2018].
```{r}
deseq_results <- lfcShrink(
  deseq_object, # The original DESeq2 object after running DESeq()
  coef = 2, # The log fold change coefficient used in DESeq(); the default is 2.
  res = deseq_results # The original DESeq2 results table
)
```

```{r}
head(deseq_results)
```


Cleaning the results to suit what we want
```{r}
# this is of class DESeqResults -- we want a data frame
deseq_df <- deseq_results %>%
  # make into data.frame
  as.data.frame() %>%
  # the gene names are row names -- let's make them a column for easy display
  tibble::rownames_to_column("Gene") %>%
  # add a column for significance threshold results
  dplyr::mutate(threshold = padj < 0.05) %>%
  # sort by statistic -- the highest values will be genes with
  # higher expression in RPL10 mutated samples
  dplyr::arrange(dplyr::desc(log2FoldChange))
```

```{r}
head(deseq_df)
```

```{r}
plotCounts(ddset, gene = "GSX1", intgroup = "Condition")
```

Creating the volcano plot
```{r}
volcano_plot <- EnhancedVolcano::EnhancedVolcano(
  deseq_df,
  lab = deseq_df$Gene,
  x = "log2FoldChange",
  y = "padj",
  pCutoff = 0.01 # Loosen the cutoff since we supplied corrected p-values
)
```

```{r}
volcano_plot
```

Creating Heat Map

- Normalizing data
```{r}
dds_norm <- vst(ddset)
```

- Choosing genes of interest
```{r}
# Calculate the variance for each gene
variances <- apply(assay(dds_norm), 1, var)

# Determine the upper quartile variance cutoff value
upper_var <- quantile(variances, 0.9995)

# Filter the data choosing only genes whose variances are in the upper quartile
df_by_var <- data.frame(assay(dds_norm)) %>%
  dplyr::filter(variances > upper_var)
```

Not-so-good Heat Map
```{r}
heatmap <- pheatmap(
  df_by_var,
  cluster_rows = TRUE, # Cluster the rows of the heatmap (genes in this case)
  cluster_cols = TRUE, # Cluster the columns of the heatmap (samples),
  show_rownames = FALSE, # There are too many genes to clearly show the labels
  main = "Heat Map",
  colorRampPalette(c(
    "deepskyblue",
    "black",
    "yellow"
  ))(25
  ),
  scale = "row" # Scale values in the direction of genes (rows)
)
```

